---
title: "Exploring Data"
output: pdf_document
date: "2023-07-07"
---

Load the necessary packages
```{r}
library(tidyverse)
library(nycflights13)
library(gapminder)
library(Lahman)
```
Load the datasets into the environment
```{r}
data("flights")
data("gapminder")
data("mpg")
```
Importing the dataset into the environment using 'dput()'
```{r}
# generates R code to recreate the data
dput(mtcars) 

# copy the R code into 'mtcars <-' 
mtcars <- structure(list(mpg = c(21, 21, 22.8, 21.4, 18.7, 18.1, 14.3, 
                       24.4, 22.8, 19.2, 17.8, 16.4, 17.3, 15.2, 10.4, 10.4, 14.7, 32.4, 
                       30.4, 33.9, 21.5, 15.5, 15.2, 13.3, 19.2, 27.3, 26, 30.4, 15.8, 
                       19.7, 15, 21.4), cyl = c(6, 6, 4, 6, 8, 6, 8, 4, 4, 6, 6, 8, 
                                                8, 8, 8, 8, 8, 4, 4, 4, 4, 8, 8, 8, 8, 4, 4, 4, 8, 6, 8, 4), 
               disp = c(160, 160, 108, 258, 360, 225, 360, 146.7, 140.8, 
                        167.6, 167.6, 275.8, 275.8, 275.8, 472, 460, 440, 78.7, 75.7, 
                        71.1, 120.1, 318, 304, 350, 400, 79, 120.3, 95.1, 351, 145, 
                        301, 121), hp = c(110, 110, 93, 110, 175, 105, 245, 62, 95, 
                                          123, 123, 180, 180, 180, 205, 215, 230, 66, 52, 65, 97, 150, 
                                          150, 245, 175, 66, 91, 113, 264, 175, 335, 109), drat = c(3.9, 
                                                                                                    3.9, 3.85, 3.08, 3.15, 2.76, 3.21, 3.69, 3.92, 3.92, 3.92, 
                                                                                                    3.07, 3.07, 3.07, 2.93, 3, 3.23, 4.08, 4.93, 4.22, 3.7, 2.76, 
                                                                                                    3.15, 3.73, 3.08, 4.08, 4.43, 3.77, 4.22, 3.62, 3.54, 4.11
                                          ), wt = c(2.62, 2.875, 2.32, 3.215, 3.44, 3.46, 3.57, 3.19, 
                                                    3.15, 3.44, 3.44, 4.07, 3.73, 3.78, 5.25, 5.424, 5.345, 2.2, 
                                                    1.615, 1.835, 2.465, 3.52, 3.435, 3.84, 3.845, 1.935, 2.14, 
                                                    1.513, 3.17, 2.77, 3.57, 2.78), qsec = c(16.46, 17.02, 18.61, 
                                                                                             19.44, 17.02, 20.22, 15.84, 20, 22.9, 18.3, 18.9, 17.4, 17.6, 
                                                                                             18, 17.98, 17.82, 17.42, 19.47, 18.52, 19.9, 20.01, 16.87, 
                                                                                             17.3, 15.41, 17.05, 18.9, 16.7, 16.9, 14.5, 15.5, 14.6, 18.6
                                                    ), vs = c(0, 0, 1, 1, 0, 1, 0, 1, 1, 1, 1, 0, 0, 0, 0, 0, 
                                                              0, 1, 1, 1, 1, 0, 0, 0, 0, 1, 0, 1, 0, 0, 0, 1), am = c(1, 
                                                                                                                      1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 
                                                                                                                      0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 1), gear = c(4, 4, 4, 3, 
                                                                                                                                                                    3, 3, 3, 4, 4, 4, 4, 3, 3, 3, 3, 3, 3, 4, 4, 4, 3, 3, 3, 
                                                                                                                                                                    3, 3, 4, 5, 5, 5, 5, 5, 4), carb = c(4, 4, 1, 1, 2, 1, 4, 
                                                                                                                                                                                                         2, 2, 4, 4, 3, 3, 3, 4, 4, 4, 1, 2, 1, 1, 2, 2, 4, 2, 1, 
                                                                                                                                                                                                         2, 2, 4, 6, 8, 2)), row.names = c("Mazda RX4", "Mazda RX4 Wag", 
                                                                                                                                                                                                                                           "Datsun 710", "Hornet 4 Drive", "Hornet Sportabout", "Valiant", 
                                                                                                                                                                                                                                           "Duster 360", "Merc 240D", "Merc 230", "Merc 280", "Merc 280C", 
                                                                                                                                                                                                                                           "Merc 450SE", "Merc 450SL", "Merc 450SLC", "Cadillac Fleetwood", 
                                                                                                                                                                                                                                           "Lincoln Continental", "Chrysler Imperial", "Fiat 128", "Honda Civic", 
                                                                                                                                                                                                                                           "Toyota Corolla", "Toyota Corona", "Dodge Challenger", "AMC Javelin", 
                                                                                                                                                                                                                                           "Camaro Z28", "Pontiac Firebird", "Fiat X1-9", "Porsche 914-2", 
                                                                                                                                                                                                                                           "Lotus Europa", "Ford Pantera L", "Ferrari Dino", "Maserati Bora", 
                                                                                                                                                                                                                                           "Volvo 142E"), class = "data.frame")


```
Create a ggplot
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y= hwy))

# Negative relationship between engine size and fuel efficiency, some points do not follow the trend, could this be related to the class of vehicle? 
```
Incorporate 'class' by mapping it to an aesthetic
```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
# The points relate to 2-seaters, these cars have larger engines but smaller bodies which improves their mileage.
```
Map 'class' to 'size aesthetic'
```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
# Size of each point corresponds to classification
# Mapping an unordered variable (class) to an ordered aesthetic (size) is not a good idea, hence the warning.
```
Map 'class' to 'alpha aesthetic'
```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```
Map 'class' to 'shape aesthetic'
```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```
Make all the points blue
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes (x = displ, y = hwy), colour = "blue")
```
Mapping a continuous variable to colour
```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, colour = year))
```
Stroke aesthetic - fills the outside of shape
```{r}
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point(shape = 21, colour = "black", fill = "white", size = 5, stroke = 5)
```
Making the aesthetic something other than a variable name
```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy, colour = displ < 5))
# sets colour based on the specified engine size, below and above 5.
```
Faceting - 1 variable
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```
Faceting - 2 variables
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl) # use facet_grid() instead of facet_wrap()
```
Faceting - not by rows or columns
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl) # use the . ~ variable to not facet by either rows or columns

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)
```
Facet a continuous variable
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~cty)
```
Geometric objects - geom_smooth()
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
Instead of setting shape of an object, you can sent linetype
```{r}
ggplot(data = mpg)+
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```
Linetype can also be substituted with colour
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    #show.legend = FALSE
  )
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```


