---
title: "Intermediate R"
output: pdf_document
date: "2023-03-15"
---
                                    Conditionals and Control Flow

== is the equality operator
!= is the inequality operator
< and > less than or greater another R object
<= and >= less than or equal to or greater than or equal to another R object
```{r}
# Comparison of logicals
TRUE == FALSE

# Comparison of numerics
-6*14 != 17-101

# Comparison of character strings
"useR" == "user"

# Compare a logical with a numeric
TRUE == 1
```
For string comparison, R determines the greater than relationship based on 
alphabetical order. Also, keep in mind that TRUE is treated as 1 for arithmetic,
and FALSE is treated as 0. Therefore, FALSE < TRUE is TRUE.
```{r}
# Comparison of numerics
-6 * 5 + 2 >= -10 + 1

# Comparison of character strings
"raining" <= "raining dogs"

# Comparison of logicals
TRUE > FALSE
```
Using relational operators, find a logical answer, i.e. TRUE or FALSE, 
for the following questions:
```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Popular days
linkedin > 15

# Quiet days
linkedin <= 5

# LinkedIn more popular than Facebook
linkedin > facebook
```
Using the relational operators you've learned so far, try to discover the 
following:
```{r}
# The social data has been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)
views <- matrix(c(linkedin, facebook), nrow = 2, byrow = TRUE)

# When does views equal 13?
views == 13

# When is views less than or equal to 14?
views <= 14
```
AND (&) operator - takes 2 logical values and returns a TRUE or FALSE
OR (|) operator - will return a TRUE or FALSE, if one comparison is correct
it will return TRUE
NOT (!) operator - negates the logical value it is used on
Double && only examines the first element of each vector - same with ||

Write R expressions to solve the following questions concerning the variable 
last:
```{r}
# The linkedin and last variable are already defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
last <- tail(linkedin, 1)
last

# Is last under 5 or above 10?
last < 5 | last > 10 

# Is last between 15 (exclusive) and 20 (inclusive)?
last > 16 & last < 20
```
Write R expressions to solve the following questions:
```{r}
# The social data (linkedin, facebook, views) has been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)
views <- matrix(c(linkedin, facebook), nrow = 2, byrow = TRUE)

# linkedin exceeds 10 but facebook below 10
linkedin > 10 & facebook < 10

# When were one or both visited at least 12 times?
linkedin >= 12 | facebook >= 12

# When is views between 11 (exclusive) and 14 (inclusive)?
views > 11 & views <= 14
```
Conditional statements

if statement
if(condition) {expression}
thre will be no print out if the expression is FALSE
```{r}
x <- -3

if(x < 0) {print("x is a negative number")}


```
else statement - has to be used together with if statement.

else will return a printout if the if codition is FALSE
```{r}
y <- 5

if(y < 0) {print("y is a negative number")
  } else {print("y is either a positive number or zero")}
  

```
else if statement - will skip if statement if it is FALSE, will return output 
if else if is TRUE, will skip else if when its FALSE
```{r}
m <- 0

if(m < 0) {print("y is a negative number")
  } else if(m == 0){print("m is zero")
  } else {print("y is either a positive number or zero")}
  
```

```{r}
# Variables related to your last day of recordings
medium <- "LinkedIn"
num_views <- 14

# Examine the if statement for medium
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
}

# Write the if statement for num_views
if (num_views > 15) {
  print("You are popular!")
}
```
Add an else

You can only use an else statement in combination with an if statement. 
The else statement does not require a condition; its corresponding code is 
simply run if all of the preceding conditions in the control structure are FALSE
```{r}
# Control structure for medium
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
} else {print("Unknown medium")}

# Control structure for num_views
if (num_views > 15) {
  print("You're popular!")
} else {print("Try to be more visible!")}
```
Customize further: else if

The else if statement allows you to further customize your control structure. 
You can add as many else if statements as you like. Keep in mind that R ignores 
the remainder of the control structure once a condition has been found that is 
TRUE and the corresponding expressions have been executed.
```{r}
# Variables related to your last day of recordings
medium <- "LinkedIn"
num_views <- 14

# Control structure for medium
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
} else if (medium == "Facebook") {print("Showing Facebook information")
  # Add code to print correct string when condition is TRUE
} else {
  print("Unknown medium")
}

# Control structure for num_views
if (num_views > 15) {
  print("You're popular!")
} else if (num_views <= 15 & num_views > 10) 
  {print("Your number of views is average")
  # Add code to print correct string when condition is TRUE
} else {
  print("Try to be more visible!")
}
```

```{r}
number <- 4
if (number < 10) {
  if (number < 5) {
    result <- "extra small"
  } else {
    result <- "small"
  }
} else if (number < 100) {
  result <- "medium"
} else {
  result <- "large"
}
print(result)
```
Go through the instructions to create R code that generates a 
'social media score', sms, based on the values of li and fb.
```{r}
# Variables related to your last day of recordings
li <- 15
fb <- 9

# If both li and fb are 15 or higher, set sms equal to double the sum 
# of li and fb.
# If both li and fb are strictly below 10, set sms equal to half the sum 
# of li and fb.
# In all other cases, set sms equal to li + fb.
# Finally, print the resulting sms variable.

# Code the control-flow construct
if (li >= 15 & fb >= 15) {
  sms <- 2 * (li + fb)
} else if (li >= 10 & fb >= 10) {
  sms <- 0.5 * (li + fb)
} else {
  sms <- li + fb
}

# Print the resulting sms to the console
sms

```
While loop, executes the code as long as the condition is TRUE

Code a while loop with the following characteristics:
The condition of the while loop should check if speed is higher than 30.
Inside the body of the while loop, print out "Slow down!".
Inside the body of the while loop, decrease the speed by 7 units and assign 
this new value to speed again. This step is crucial; otherwise your while loop 
will never stop and your session will expire.
```{r}
# Initialize the speed variable
speed <- 64

# Code the while loop
while (speed > 30) {
  print(paste("Slow down!"))
  speed <- speed - 7
}

# Print out the speed variable
speed
```
If the speed is greater than 48, have R print out "Slow down big time!", 
and decrease the speed by 11.
Otherwise, have R simply print out "Slow down!", and decrease the speed by 6.
```{r}
# Initialize the speed variable
speed <- 64

# Extend/adapt the while loop
while (speed > 30) {
  print(paste("Your speed is",speed))
  if ( speed > 48) {
    print(paste("Slow down big time!"))
    speed <- speed - 11
  } else {
    print(paste("Slow down!"))
    speed <- speed - 6
  }
}
```
Adapt the while loop such that it is abandoned when the speed of the 
vehicle is greater than 80. This time, the speed variable has been initialized 
to 88; keep it that way.
```{r}
# Initialize the speed variable
speed <- 88

while (speed > 30) {
  print(paste("Your speed is", speed))
  
  # Break the while loop when speed exceeds 80
  if (speed > 80) {
    break
  }
  
  if (speed > 48) {
    print("Slow down big time!")
    speed <- speed - 11
  } else {
    print("Slow down!")
    speed <- speed - 6
  }
}
```
Finish the while loop so that it:

prints out the triple of i, so 3 * i, at each run.
is abandoned with a break if the triple of i is divisible by 8, 
but still prints out this triple before breaking.
```{r}
# Initialize i as 1 
i <- 1

# Code the while loop
while (i <= 10) {
  print(3 * i)
  if ( (3 * i) %% 8 == 0) {
    break
  }
  i <- i + 1
}
```
Loop over a vector

In the previous video, Filip told you about two different strategies for 
using the for loop. To refresh your memory, consider the following loops that 
are equivalent in R:
```{r}
primes <- c(2, 3, 5, 7, 11, 13)

# loop version 1
for (p in primes) {
  print(p)
}

# loop version 2
for (i in 1:length(primes)) {
  print(primes[i])
}
```
Write a for loop that iterates over all the elements of linkedin and prints 
out every element separately. Do this in two ways: using the loop version 1 and 
the loop version 2 in the example code above.
```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Loop version 1
for(v in linkedin) {
    print(v)
}

# Loop version 2
for (v in 1:length(linkedin)) {
  print(linkedin[v])
}
```
Loop over a list

Looping over a list is just as easy and convenient as looping over a vector. 
There are again two different approaches here:
```{r}
primes_list <- list(2, 3, 5, 7, 11, 13)

# loop version 1
for (p in primes_list) {
  print(p)
}

# loop version 2
for (i in 1:length(primes_list)) {
  print(primes_list[[i]])
}
```
Loop directly over the nyc list (loop version 1).
Define a looping index and do subsetting using double brackets (loop version 2).
```{r}
# The nyc list is already specified
nyc <- list(pop = 8405837, 
            boroughs = c("Manhattan", "Bronx", "Brooklyn", 
                         "Queens", "Staten Island"), 
            capital = FALSE)

# Loop version 1
for (p in nyc) {
  print(p)
}

# Loop version 2
for (p in 1:length(nyc)) {
  print(nyc[[p]])
}
```
Add code to the for loop that loops over the elements of the linkedin vector:

If the vector element's value exceeds 10, print out "You're popular!".
If the vector element's value does not exceed 10, print out "Be more visible!"
```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Code the for loop with conditionals
for (li in linkedin) {
  if (li > 10) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  print(li)
}
```
Extend the for loop with two new, separate if tests as follows:

If the vector element's value exceeds 16, print out "This is ridiculous, 
I'm outta here!" and have R abandon the for loop (break).
If the value is lower than 5, print out "This is too embarrassing!" 
and fast-forward to the next iteration (next).
```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Adapt/extend the for loop
for (li in linkedin) {
  if (li > 10) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  
  # Add if statement with break
  if (li > 16) {
    print("This is ridiculous, I'm outta here!")
      break
  }
  
  # Add if statement with next
  if (li < 5) {
    print("This is too embarrassing!") 
      next
  }
  
  print(li)
}
```
Build a for loop from scratch

This exercise will not introduce any new concepts on for loops.

We already went ahead and defined a variable rquote. 
This variable has been split up into a vector that contains separate letters 
and has been stored in a vector chars with the strsplit() function.

Can you write code that counts the number of r's that come before 
the first u in rquote?
```{r}
# Pre-defined variables
rquote <- "r's internals are irrefutably intriguing"
chars <- strsplit(rquote, split = "")[[1]]

# Initialize rcount
rcount <- 0

# Finish the for loop
for (char in chars) {
    if (char == "r")
    rcount <- rcount + 1
    if (char == "u")
    break
  
  
}

# Print out rcount
rcount
```
                                    FUNCTIONS
Look up the documentation for an r function either using help() or ?
e.g. help(sd) or ?sd

can also use args() to learn about the arguments of a function instead of 
reading the documentation

```{r}
help(sd)
?sd
args(sd)
```
In the next exercises, you'll be learning how to use the mean() function with 
increasing complexity. The first thing you'll have to do is get acquainted with
the mean() function.
```{r}
# Consult the documentation on the mean() function
?mean

# Inspect the arguments of the mean() function
args(mean)
```
Once more, you'll be working with the view counts of your social network 
profiles for the past 7 days. These are stored in the linkedin and facebook 
vectors and have already been created for you.
```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Calculate average number of views
avg_li <- mean(linkedin)
avg_fb <- mean(facebook)


# Inspect avg_li and avg_fb
avg_li
avg_fb
```

```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Calculate the mean of the sum
avg_sum <- mean(linkedin + facebook)

# Calculate the trimmed mean of the sum
avg_sum_trimmed <- mean(linkedin + facebook, trim = 0.2)

# Inspect both new variables
avg_sum
avg_sum_trimmed

# When the trim argument is not zero, it chops off a fraction (equal to trim) 
# of the vector you pass as argument x.
```
```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, NA, 17, 14)
facebook <- c(17, NA, 5, 16, 8, 13, 14)

# Basic average of linkedin
mean(linkedin)

# Advanced average of linkedin
mean(linkedin, na.rm = TRUE)
```
Use abs() on linkedin - facebook to get the absolute differences between 
the daily LinkedIn and Facebook profile views. Place the call to abs() inside 
mean() to calculate the Mean Absolute Deviation. In the mean() call, 
make sure to specify na.rm to treat missing values correctly!
```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, NA, 17, 14)
facebook <- c(17, NA, 5, 16, 8, 13, 14)

# Calculate the mean absolute deviation
mean(abs(linkedin - facebook), na.rm = TRUE)
```
                                  Write your own function
```{r}
# Create a function pow_two()
pow_two <- function(x) {
    x ** 2
}

# Use the function
pow_two(12)

# Create a function sum_abs()
sum_abs <- function(x,y) {
    abs(x) + abs(y)
}

# Use the function
sum_abs(-2, 3)
```
Let's say you want to write a function that gives us the random outcome of 
throwing a fair die:
```{r}
throw_die <- function() {
  number <- sample(1:6, size = 1)
  number
}

throw_die()
```
Define a function, hello(). It prints out "Hi there!" and returns TRUE. 
It has no arguments.
Call the function hello(), without specifying arguments of course.
```{r}
# Define the function hello()
hello <- function() {
    print("Hi there!")
    TRUE
}

# Call the function hello()
hello()
```
Add an optional argument, named print_info, that is TRUE by default.
Wrap an if construct around the print() function: 
this function should only be executed if print_info is TRUE.
Feel free to experiment with the pow_two() function you've just coded.
```{r}
pow_two <- function(x, print_info = TRUE) {
  y <- x ^ 2
  if (print_info == TRUE)
  print(paste(x, "to the power two equals", y))
  return(y)
}

pow_two(5)
```
Finish the function definition for interpret(), that interprets the number of 
profile views on a single day:
  The function takes one argument, num_views.
  If num_views is greater than 15, the function prints out "You're popular!" 
  to the console and returns num_views.
  Else, the function prints out "Try to be more visible!" and returns 0.
  Finally, call the interpret() function twice: on the first value of the 
  linkedin vector and on the second element of the facebook vector.
```{r}
# Define the interpret function
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}

# Call the interpret function
interpret(linkedin[1])
```
Finish the template for the interpret_all() function:

  Make return_sum an optional argument, that is TRUE by default.
  Inside the for loop, iterate over all views: on every iteration,
  add the result of interpret(v) to count. Remember that interpret(v) returns v 
  for popular days, and 0 otherwise. At the same time, interpret(v) will also do 
  some printouts.
Finish the if construct:
  If return_sum is TRUE, return count.
  Else, return NULL.
Call this newly defined function on both linkedin and facebook.
```{r}
# The interpret() can be used inside interpret_all()
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}

# Define the interpret_all() function
# views: vector with data to interpret
# return_sum: return total number of views on popular days?
interpret_all <- function(views, return_sum = TRUE) {
  count <- 0

  for (v in views) {
    count <- interpret(v) + count
  }
  if (return_sum) {
    TRUE
    return(count)
  } else {
    return(NULL)
  }
}

# Call the interpret_all() function 
interpret_all(linkedin)
```
                                  The Apply Family
Usually follows this expression - lapply(X, FUN, ...)
lapplu is used to apply a function over each element of a list or vector
```{r}
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")

# Split names from birth year
split_math <- strsplit(pioneers, split = ":")

# Convert to lowercase strings: split_low
split_low <- lapply(split_math, tolower)

# Take a look at the structure of split_low
str(split_low)
```
Apply select_first() over the elements of split_low with lapply() and assign 
the result to a new variable names.
Next, write a function select_second() that does the exact same thing for 
the second element of an inputted vector.
Finally, apply the select_second() function over split_low and assign the 
output to the variable years.
```{r}
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split <- strsplit(pioneers, split = ":")
split_low <- lapply(split, tolower)

# Write function select_first()
select_first <- function(x) {
  x[1]
}

# Apply select_first() over split_low: names
names <- lapply(split_low, select_first)

# Write function select_second()
select_second <- function(x) {
  x[2]
}

# Apply select_second() over split_low: years
years <- lapply(split_low, select_second)
```
Anonymous functions
```{r}
# Transform: use anonymous function inside lapply
names <- lapply(split_low, function(x) {x[1]})

# Transform: use anonymous function inside lapply
years <- lapply(split_low, function(x) {x[2]})
```
Use lapply() twice to call select_el() over all elements in split_low: 
once with the index equal to 1 and a second time with the index equal to 2. 
Assign the result to names and years, respectively.
```{r}
# Generic select function
select_el <- function(x, index) {
  x[index]
}

# Use lapply() twice on split_low: names and years
names <- lapply(split_low, select_el, index = 1)
years <- lapply(split_low, select_el, index = 2)
```
sapply():
Short-hand for simplify apply
Follows the same structure - sapply(X, FUN, ...)
```{r}
temp <- list(c(3, 7, 9, 6, -1), c(6, 9, 12, 13, 5), c(4, 8, 3, -1, -3
), c(1, 4, 7, 2, -2), c(5, 7, 9, 4, 2), c(-3, 5, 8, 9, 4), c(3, 
6, 9, 4, 1))

# Use lapply() to find each day's minimum temperature
lapply(temp, min)

# Use sapply() to find each day's minimum temperature
sapply(temp, min)

# Use lapply() to find each day's maximum temperature
lapply(temp, max)

# Use sapply() to find each day's maximum temperature
sapply(temp, max)
```

```{r}
# Function calculates the average of the min and max temperatures of temp
extremes_avg <- function(x) {
  (min(x) + max(x)) / 2
}

# Apply extremes_avg() over temp using sapply()
sapply(temp, extremes_avg)

# Apply extremes_avg() over temp using lapply()
lapply(temp, extremes_avg)
```

```{r}
# A function that returns min and max of a vector: extremes
extremes <- function(x) {
  c(min = min(x), max = max(x))
}

# Apply extremes() over temp with sapply()
sapply(temp, extremes)

# Apply extremes() over temp with lapply()
lapply(temp, extremes)
```
vapply():
Syntax -> vapply(X, FUN, FUN.VALUE, ..., USE.NAMES = TRUE)

vapply() can be considered a more robust version of sapply(), because you 
explicitly restrict the output of the function you want to apply. Converting 
your sapply() expressions in your own R scripts to vapply() expressions is 
therefore a good practice.
```{r}
# Definition of basics()
basics <- function(x) {
  c(min = min(x), mean = mean(x), max = max(x))
}

# Apply basics() over temp using vapply()
vapply(temp, basics, numeric(3)) # there are 3 vectors in the function
```
Increase the vector length to 4 by adjusting the function
```{r}
# Definition of the basics() function
basics <- function(x) {
  c(min = min(x), mean = mean(x), median = median(x), max = max(x))
}
# Apply basics() over temp using vapply()
vapply(temp, basics, numeric(4))
```
                                  Utilities
Useful functions
abs(): Calculate the absolute value.
sum(): Calculate the sum of all the values in a data structure.
mean(): Calculate the arithmetic mean.
round(): Round the values to 0 decimal places by default.
```{r}
# The errors vector has already been defined for you
errors <- c(1.9, -2.6, 4.0, -9.5, -3.4, 7.3)

# Sum of absolute rounded values of errors
sum(round(abs(errors)))
```
Using mean()
```{r}
# New vectors
vec1 <- c(1.5, 2.5, 8.4, 3.7, 6.3)
vec2 <- rev(vec1)

# Find the mean of vec1 and vec2
mean(c(abs(vec1), abs(vec2)))
```                          
R features a bunch of functions to juggle around with data structures:
seq(): Generate sequences, by specifying the from, to, and by arguments.
rep(): Replicate elements of vectors and lists.
sort(): Sort a vector in ascending order. Works on numerics, but also on 
character strings and logicals.
rev(): Reverse the elements in a data structures for which reversal is defined.
str(): Display the structure of any R object.
append(): Merge vectors or lists.
is.*(): Check for the class of an R object.
as.*(): Convert an R object from one class to another.
unlist(): Flatten (possibly embedded) lists to produce a vector.
```{r}
# The linkedin and facebook lists have already been created for you
linkedin <- list(16, 9, 13, 5, 2, 17, 14)
facebook <- list(17, 7, 5, 16, 8, 13, 14)

# Convert linkedin and facebook to a vector: li_vec and fb_vec
li_vec <- unlist(linkedin)
fb_vec <- unlist(facebook)

# Append fb_vec to li_vec: social_vec
social_vec <- append(li_vec, fb_vec)

# Sort social_vec
sort(social_vec, decreasing = TRUE)
```
rep goes before seq
```{r}
rep(seq(1, 7, by = 2), times = 7)
```
Creating sequences using seq()
```{r}
# Create first sequence: seq1
seq1 <- seq(1, 500, by = 3)

# Create second sequence: seq2
seq2 <- seq(1200, 900, by = -7)

# Calculate total sum of the sequences
sum(seq1) + sum(seq2)
```
Regular expressions an be used to see whether a pattern exists inside a 
character string or a vector of character strings. For this purpose, 
you can use:
grepl(pattern = <regex>, x = <string>)
grep(pattern = <regex>, x = <string>)

With sub() and gsub() you can specify a replacement argument: 
sub(pattern = <regex>, replacement = <str>, x = <str>)
gsub(pattern = <regex>, replacement = <str>, x = <str>)
```{r}
# The emails vector
emails <- c("john.doe@ivyleague.edu", "education@world.gov", "dalai.lama@peace.org",
            "invalid.edu", "quant@bigdatacollege.edu", "cookie.monster@sesame.tv")

# Use grepl() to match for "edu"
grepl(pattern = "edu", x = emails)

# Use grep() to match for "edu", save result to hits
hits <- grep(pattern = "edu", x = emails)
hits

# Subset emails using hits
emails[hits]
```
You can use the caret, ^, and the dollar sign, $ to match the content located in the start and end of a string, respectively. Other regular expressions:

.*: A usual suspect! It can be read as "any character that is matched zero or 
more times".

\\s: Match a space. The "s" is normally a character, escaping it (\\) makes it 
a metacharacter.

[0-9]+: Match the numbers 0 to 9, at least once (+).

([0-9]+): The parentheses are used to make parts of the matching string
available to define the replacement. 

```{r}
# Use grepl() to match for .edu addresses more robustly
grepl(pattern = "@.*\\.edu$", x = emails)

# Use grep() to match for .edu addresses more robustly, save result to hits
hits <- grep(pattern = "@.*\\.edu$", x = emails)

# Subset emails using hits
emails[hits]
```
With the advanced regular expression "@.*\\.edu$", use sub() to replace the 
match with "@datacamp.edu". Since there will only be one match per character 
string, gsub() is not necessary here. Inspect the resulting output.
```{r}
# Use sub() to convert the email domains to datacamp.edu
sub(pattern = "@.*\\.edu$", replacement = "@datacamp.edu", x = emails)
```
Dates and Times:
The character string has to obey a format that can be defined using a set of 
symbols (the examples correspond to 13 January, 1982):

%Y: 4-digit year (1982)
%y: 2-digit year (82)
%m: 2-digit month (01)
%d: 2-digit day of the month (13)
%A: weekday (Wednesday)
%a: abbreviated weekday (Wed)
%B: month (January)
%b: abbreviated month (Jan)
```{r}
# Get the current date: today
today <- Sys.Date()
today

# See what today looks like under the hood
unclass(today)

# Get the current time: now
now <- Sys.time()
now

# See what now looks like under the hood
unclass(now)
```
To create a Date object from a simple character string in R, you can use the 
as.Date()
```{r}
# Definition of character strings representing dates
str1 <- "May 23, '96"
str2 <- "2012-03-15"
str3 <- "30/January/2006"

# Convert the strings to dates: date1, date2, date3
date1 <- as.Date(str1, format = "%b %d, '%y")
date1

date2 <- as.Date(str2)
date2 

date3 <- as.Date(str3, format = "%d/%B/%Y")
date3 

# Convert dates to formatted strings
format(date1, "%A") # Select weekday
format(date2, "%d") # Select day of the month
format(date3, "%b %Y") # Select abbreviated month and 4-digit year
```
Create and format times
Similar to working with dates, you can use as.POSIXct() to convert from a 
character string to a POSIXct object, and format() to convert from a POSIXct 
object to a character string. Again, you have a wide variety of symbols:

%H: hours as a decimal number (00-23)
%I: hours as a decimal number (01-12)
%M: minutes as a decimal number
%S: seconds as a decimal number
%T: shorthand notation for the typical format %H:%M:%S
%p: AM/PM indicator
```{r}
# Definition of character strings representing times
str1 <- "May 23, '96 hours:23 minutes:01 seconds:45"
str2 <- "2012-3-12 14:23:08"

# Convert the strings to POSIXct objects: time1, time2
time1 <- as.POSIXct(str1, format = "%B %d, '%y hours:%H minutes:%M seconds:%S")
time1

time2 <- as.POSIXct(str2)
time2

# Convert times to formatted strings
format(time1, "%M") # Extract only minutes
format(time2, "%I:%M %p") # Extract hours:minutes AM/PM
```
Calculations with Dates:
```{r}
# Create some dates for when I ate pizza
day1 <- structure(19532, class = "Date") # Equivalent as.Date("2023-06-24")
day2 <- structure(19534, class = "Date")
day3 <- structure(19539, class = "Date")
day4 <- structure(19545, class = "Date")
day5 <- structure(19550, class = "Date")

pizza <- c(day1, day2, day3, day4, day5)

# Difference between last and first pizza day
day5 - day1

# Create differences between consecutive pizza days: day_diff
day_diff <- diff(pizza)
day_diff

# Average period between two consecutive pizza days
mean(day_diff)
```
Calculations with Times:
```{r}
# Create a vector for the dates and times a user has logged in
login <- structure(c(1687947484.74737, 1688375658.36921, 1688386911.68576, 
1688387844.83761, 1688593075.89139), class = c("POSIXct", "POSIXt"
))
# Create a vector for the dates and times a user has logged out
logout <- structure(c(1687947484.74737, 1688375658.36921, 1688386911.68576, 
1688387844.83761, 1688593075.89139), class = c("POSIXct", "POSIXt"
))

# Calculate the difference between login and logout: time_online
time_online <- logout - login

# Inspect the variable time_online
print(time_online)

# Calculate the total time online
sum(time_online)

# Calculate the average time online
mean(time_online)
```
Use as.Date() to convert the astro vector to a vector containing Date objects. 
You will need the %d, %b and %Y symbols to specify the format. 
Store the resulting vector as astro_dates.

Use as.Date() to convert the meteo vector to a vector with Date objects. 
This time, you will need the %B, %d and %y symbols for the format argument. 
Store the resulting vector as meteo_dates.

With a combination of max(), abs() and -, calculate the maximum absolute 
difference between the astronomical and the meteorological beginnings of a 
season, i.e. astro_dates and meteo_dates. Simply print this maximum difference 
to the console output.
```{r}
astro <- c(spring = "20-Mar-2015", summer = "25-Jun-2015", fall = "23-Sep-2015", 
winter = "22-Dec-2015")

meteo <- c(spring = "March 1, 15", summer = "June 1, 15", fall = "September 1, 15", 
winter = "December 1, 15")

# Convert astro to vector of Date objects: astro_dates
astro_dates <- as.Date(astro, format = "%d-%b-%Y")
astro_dates

# Convert meteo to vector of Date objects: meteo_dates
meteo_dates <- as.Date(meteo, format = "%B %d, %y")
meteo_dates

# Calculate the maximum absolute difference between astro_dates and meteo_dates
max(abs(astro_dates - meteo_dates))
```

