---
Title: Intro to R
Output: pdf_document
Date: "2023-03-13"
output:
  pdf_document: default
  word_document: default
---

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Calculate total gains for poker and roulette
total_poker <-sum(poker_vector)
total_roulette <-sum(roulette_vector)

# Check if you realized higher total gains in poker than in roulette
total_poker > total_roulette
```
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Select poker results for Monday, Tuesday and Wednesday
poker_start <- poker_vector[c("Monday", "Tuesday", "Wednesday")]
  
# Calculate the average of the elements in poker_start
mean(poker_start)
```
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- poker_vector > 0
  
# Print out selection_vector
selection_vector
```
```{r}
# Box office Star Wars (in millions!)
new_hope <- c(460.998, 314.4)
empire_strikes <- c(290.475, 247.900)
return_jedi <- c(309.306, 165.8)

# Create box_office
box_office <- c(new_hope, empire_strikes, return_jedi)

# Construct star_wars_matrix
star_wars_matrix <- matrix(box_office, nrow = 3, byrow = TRUE)

star_wars_matrix
```
```{r}
# Box office Star Wars (in millions!)
new_hope <- c(460.998, 314.4)
empire_strikes <- c(290.475, 247.900)
return_jedi <- c(309.306, 165.8)

# Construct matrix
star_wars_matrix <- matrix(c(new_hope, empire_strikes, return_jedi), 
                           nrow = 3, byrow = TRUE)

# Vectors region and titles, used for naming
region <- c("US", "non-US")
titles <- c("A New Hope", "The Empire Strikes Back", "Return of the Jedi")

# Name the columns with region
colnames(star_wars_matrix) <- region

# Name the rows with titles
rownames(star_wars_matrix) <- titles

# Print out star_wars_matrix
star_wars_matrix
```
                        FACTORS
Factors in R are used to store categorical variables. A categorical variable 
can belong to a limited number of categories unlike a continuous variable. 
A good example of a categorical variable is 'sex' (Male or Female)
```{r}
# Assign to the variable theory what this chapter is about!
theory <- "factors"
```
To create factors in R, you make use of the function factor().

sex_vector <- c("Male","Female","Female","Male","Male")
It is clear that there are two categories, or in R-terms 'factor levels', 
at work here: "Male" and "Female".

The function factor() will encode the vector as a factor
```{r}
# Sex vector
sex_vector <- c("Male", "Female", "Female", "Male", "Male")

# Convert sex_vector to a factor
factor_sex_vector <- factor(sex_vector)

# Print out factor_sex_vector
factor_sex_vector
```
2 types of categorical variables are nominal and ordinal. 

Nominal means there is no implied order e.g. Elephant, Giraffe, Donkey etc.

Ordinal have natural ordering e.g. Low, Medium, High.
```{r}
# Animals
animals_vector <- c("Elephant", "Giraffe", "Donkey", "Horse")
factor_animals_vector <- factor(animals_vector)
factor_animals_vector

# Temperature
temperature_vector <- c("High", "Low", "High","Low", "Medium")
factor_temperature_vector <- factor(temperature_vector, order = TRUE, 
                                    levels = c("Low", "Medium", "High"))
factor_temperature_vector
```
```{r}
# Code to build factor_survey_vector
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
factor_survey_vector

# Specify the levels of factor_survey_vector
levels(factor_survey_vector) <- c("Female", "Male")
factor_survey_vector
```
summary() function  will give you a quick overview of the contents of a 
variable e.g. how many "Female" and "Male" responses there were in a survey.

summary(my_var)
```{r}
# Build factor_survey_vector with clean levels
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
levels(factor_survey_vector) <- c("Female", "Male")
factor_survey_vector

# Generate summary for survey_vector
summary(survey_vector)

# Generate summary for factor_survey_vector
summary(factor_survey_vector)
```
Sometimes you will also deal with factors that do have a natural ordering 
between its categories. If this is the case, we have to make sure that we 
pass this information to R.
```{r}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
```
By default, the function factor() transforms speed_vector into an unordered 
factor. To create an ordered factor, you have to add two additional arguments: 
ordered and levels.
```{r}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")

# Convert speed_vector to ordered factor vector
factor_speed_vector <-factor(speed_vector,
                            ordered = TRUE,
                            levels = c("slow", "medium",
                                        "fast"))

# Print factor_speed_vector
factor_speed_vector
summary(factor_speed_vector)
```
```{r}
# Create factor_speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
factor_speed_vector <- factor(speed_vector, ordered = TRUE, 
                              levels = c("slow", "medium", "fast"))
factor_speed_vector
# Factor value for second data analyst
da2 <- factor_speed_vector[2]
da2
# Factor value for fifth data analyst
da5 <- factor_speed_vector[5]
da5
# Is data analyst 2 faster than data analyst 5?
da2 > da5
```
                                          DATAFRAMES
These contain different data types instead of only one.
A data frame has the variables of a dataset as columns and the observations 
as rows. 
```{r}
# Print out built-in R data frame
mtcars 
```
The function head() enables you to show the first observations of a data frame. 
Similarly, the function tail() prints out the last observations in your dataset.
```{r}
# Call head() on mtcars
head(mtcars)
```
Another method that is often used to get a rapid overview of your data is the 
function str(). The function str() shows you the structure of your data set.
```{r}
# Investigate the structure of mtcars
str(mtcars)
```
You construct a data frame with the data.frame() function. 
As arguments, you pass the vectors from before: they will become the different 
columns of your data frame.
```{r}
# Definition of vectors
name <- c("Mercury", "Venus", "Earth", 
          "Mars", "Jupiter", "Saturn", 
          "Uranus", "Neptune")
type <- c("Terrestrial planet", 
          "Terrestrial planet", 
          "Terrestrial planet", 
          "Terrestrial planet", "Gas giant", 
          "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 
              11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03, 
              0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

# Create a data frame from the vectors
planets_df <- data.frame(name, type, diameter, rotation, rings)
planets_df
```

```{r}
# Check the structure of planets_df
str(planets_df)
```
Selection of data frame elements.
my_df[1,2] selects the value at the first row and second column.
my_df[1:3,2:4] selects rows 1, 2, 3 and columns 2, 3, 4.
my_df[1, ] selects all elements of the first row.
```{r}
# Print out diameter of Mercury (row 1, column 3)
planets_df[1,3]

# Print out data for Mars (entire fourth row)
planets_df[4,]
```
Instead of using numerics to select elements of a data frame, you can also 
use the variable names to select columns of a data frame.
```{r}
# Select first 5 values of diameter column
planets_df[1:5, "diameter"]
```
You will often want to select an entire column, namely one specific variable 
from a data frame.

There is a short-cut. If your columns have names, you can use the $ sign.
```{r}
# Select the rings variable from planets_df
rings_vector <- planets_df$rings
  
# Print out rings_vector
rings_vector
```
The code in the editor selects the name column of all planets that have rings. 
Adapt the code so that instead of only the name column, all columns for planets 
that have rings are selected.

Remember that to select all columns, you simply have to leave the columns part 
inside the [ ] empty! This means you'll need [rings_vector, ]
```{r}
# Adapt the code to select all columns for planets with rings
planets_df[rings_vector, "name"]

# Adapt the code to select all columns for planets with rings
planets_df[rings_vector, ]
```
You should see the subset() function as a short-cut to do exactly the same as 
what you did in the previous exercises.

The code below will give the exact same result as you got in the previous 
exercise, but this time, you didn't need the rings_vector!
```{r}
# Same as before but without the rings_vector
subset(planets_df, subset = rings)

# Select planets with diameter < 1
subset(planets_df, subset = diameter <1)
```
In data analysis you can sort your data according to a certain variable in 
the dataset. In R, this is done with the help of the function order().
```{r}
# Play around with the order function in the console
m <- c(30, 3, 3000, 300, 30000, 3000000, 300000)
m

order(m)

m[order(m)]
```
You would like to rearrange your data frame such that it starts with the 
smallest planet and ends with the largest one. A sort on the diameter column.
```{r}
# planets_df is pre-loaded in your workspace
planets_df

# Use order() to create positions
positions <-  order(planets_df$diameter)
positions

# Use positions to sort planets_df
planets_df[positions, ]
```
                                            LISTS
A list in R allows you to gather a variety of objects under one name 
(that is, the name of the list) in an ordered way. These objects can be 
matrices, vectors, data frames, even other lists, etc. It is not even 
required that these objects are related to each other in any way.

To construct a list you use the function list()
```{r}
# Vector with numerics from 1 up to 10
my_vector <- 1:10 

# Matrix with numerics from 1 up to 9
my_matrix <- matrix(1:9, ncol = 3)

# First 10 elements of the built-in data frame mtcars
my_df <- mtcars[1:10,]

# Construct list with these different elements:
my_list <- list(my_vector, my_matrix, my_df)
my_list
```

```{r}
# Adapt list() call to give the components names
my_list <- list(vec = my_vector, 
                mat = my_matrix, 
                df = my_df)

# Print out my_list
my_list
```

